---
title: "Automation, Scheduling and Debugging"
subtitle: ""
author: "Carol Sobral, Linus Hagemann and Laia Domenech Burin"
date: "(Fall 2026) Introduction to Data Science"
output: 
    rmdformats::robobook:
    toc: TRUE
    df_print: paged
    number_sections: FALSE
    highlight: tango
    toc_depth: 3
    toc_float: true
    self_contained: false
---

```{=html}
<style>
.h1,h2,h3 {
color:#2f1a61;
}

.subtitle, section.normal {
color:#291854;
}

.title {
color:#cc0065;
}

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Welcome to Week 3! 👋


This week's tutorial will be divided in two parts. 

1.  More Iteration practice
2.  Automation with R scripts 
3.  Scheduling with `cronR` and `taskscheduleR`

---


# 1. More Iteration with `purrr` 🔁

We saw some examples of iteration with `purrr::map()` in the last session. Now that we have also covered it in the Lectures let's do a few more examples to get more comfortable with this approach.

## Quick Recap

Remember the `map()` family syntax:
```r
map(.x, .f, ...)
```
-   `purrr::map()` returns a list.
-   `purrr::map_lgl()` returns a logical vector
-   `purrr::map_int()` returns an integer vector.
-   `purrr::map_dbl()` returns a double vector.
-   `purrr::map_chr()` returns a character vector.

---

## Introduction to Nesting

Sometimes we want to perform the same analysis on different groups in our data. For example, we might want to calculate summary statistics for each species of penguin, or fit a model for each country in a dataset.

**Nesting** is a powerful technique that helps us organize grouped data. When we nest data:

1. We group our data by some variable(s)
2. All the remaining data for each group gets "packed" into a special list-column
3. Each row now represents one group, with all that group's data stored in the list-column

```r
# Before nesting: regular dataframe
penguins
#> species    bill_length_mm  bill_depth_mm  ...
#> Adelie     39.1           18.7           ...  
#> Adelie     39.5           17.4           ...
#> Chinstrap  46.5           17.9           ...

# After nesting by species: each species gets one row
penguins %>% 
  group_by(species) %>% 
  nest()

#> species    data              
#> Adelie     <tibble [152 × 7]>
#> Chinstrap  <tibble [68 × 7]> 
#> Gentoo     <tibble [124 × 7]>
```

The magic happens when we combine nesting with `purrr::map()` - we can apply the same function to each group's data, then unnest the results to get a clean summary table.

## Exercise: Penguin Flipper analysis

Let's look at our penguin dataset again as an example.

```{r}
library(palmerpenguins)
library(tidyverse)
data(penguins)
```

:::alert-info
**Exercise 1**
Group the penguins by species and nest the remaining data into a column called data. Save the result in a new dataframe called `penguins_nested`:

```{r, echo= FALSE, eval= FALSE}
# Your code here: Use group_by() and nest()
```
:::


:::alert-info
**Exercise 2**
Write a function called `summarize_bills` that takes a dataframe and returns:
- `count`: number of rows
- `mean_bill_length`: mean bill length (handle NAs)
- `max_bill_length`: max bill length (handle NAs)

```{r, echo= FALSE, eval= FALSE}
summarize_bills <- function(df) {
  # Your code here: use summarize() to create the three variables
}

# Test your function on the first species
summarize_bills(penguins_nested$data[[1]])
```
:::


:::alert-info
**Exercise 3**
Add a new column called `bill_stats` to `penguins_nested` that applies your `summarize_bills` function to each nested dataframe using `map()`:

```{r, echo= FALSE, eval= FALSE}
penguins_nested <- penguins_nested %>%
  # Your code here: mutate() with map()

# Look at the result
penguins_nested
```

:::


:::alert-info
**Exercise 4**
Create a clean summary table by selecting just `species` and `bill_stats`, then unnesting:

```{r, echo= FALSE, eval= FALSE}
penguins_summary <- penguins_nested %>%
  # Your code here: select() and unnest()

penguins_summary
```
:::


:::alert-info
**Exercise 5**

Write the complete pipeline from the original `penguins` data to the final summary in one go:

```{r, echo= FALSE, eval= FALSE}
final_summary <- penguins %>%
  # Your complete pipeline here

final_summary
```
:::

--- 

# 2. Automation of Projects 🤖

When working on a data science project, you will often have to repeat the same steps multiple times. For example, you might have to:

- import data
- clean data
- run analyses
- create visualizations
- generate reports
- etc.

To avoid doing these steps manually every time, you can automate them using pipelines. We already saw pipelines in the context of the `dplyr` package using the pipe operator (`%>%` or `|>`), now we will look at how to automate entire workflows of R scripts using the `source()` function.


## Folder structure and thinking in pipelines ⚙️

When working on a data science project, it is important to keep your files organized. An example folder structure based on the example you saw in the lecture might look as follows:
```
lotr_project/
├── data/
│   ├── raw/
│   └── processed/
├── scripts/
│   ├── Makefile.R
│   └── 00-packages.R
│   └── 01-download-data.R
│   └── 02-process-data.R
│   └── 03-plot.R
├── outputs/
│   ├── figures/
│   └── tables/
└── README.md
```

In this example, the `data/` folder contains raw and processed data, the `scripts/` folder contains R scripts that perform specific tasks, and the `outputs/` folder contains figures and tables generated by the scripts. The `Makefile.R` script is the main script that sources the other scripts in the correct order.

---

## Let's try it ourselves

To work with this example I have created all the files and folders as above in this week's lab within the lotr_project folder. Try running the `Makefile.R` script to see how it works. You can open the different scripts and run the makefile line by line to see what each part does.

:::alert-info
**Exercise 1**
Can you add a new script that creates a summary table of the words spoken by Species and Film and saves it to the `outputs/tables/` folder? You can use the `dplyr` package to create the summary table and the `write_tsv()` function to save it as a .tsv file. Don't forget to source your new script in the `Makefile.R` script.
:::

:::alert-info
**Exercise 2**
Update the `Makefile.R` script so that it also clears out the table you created in the exercise above.
:::

:::alert-info
**Exercise 3**
Do your new script and the 03-plot.R script need to be run sequentially? How could we think about these two scripts in terms of the pipeline graphs you saw in the lecture?
:::

---

# 3. Scheduling with `cronR` and `taskscheduleR` 📅
Now that we have a working pipeline, let's learn how to schedule it to run automatically!

Task scheduling allows you to:

- Run your pipeline at specific times (daily, weekly, monthly)
- Process new data automatically
- Generate reports on a schedule
- Perform maintenance tasks

### Aside: Relative vs Absolute Paths 🗺️

When working with file paths in R, you can use either relative or absolute paths. So far we have used relative paths as this makes it easier to share code with others. When we schedule tasks however the working directory might not be what we expect, so it is often safer to use absolute paths. A useful package for this is `here` which allows you to construct file paths relative to the root of your project. You will see some examples of this in the code below. (Note that for this reason also, the lotr project as we have set it up above will not work if you just run the `Makefile.R` script, as it uses relative paths. You would have to amend the script to use absulute paths, the `here` package or set your working directory to the lotr_project folder first. Below it is inlcuded just for demonstration.)


## Scheduling with `cronR` on Mac and Linux ⏰

```{r eval = FALSE}
# install.packages("cronR")
# install.packages("here")
library(cronR)

# Define the path to your Master R script
cmd <- cron_rscript(here::here("03_automation","lotr_project","scripts","Makefile.R"))

# Add a new cron job to run the script every day at 1am
cron_add(
  command = cmd,
  frequency = "daily",
  at = "01:00",
  id = "lotr_pipeline_daily",
  description = "Daily LOTR analysis pipeline"
)

# List all scheduled tasks
cron_ls()

# To remove a scheduled task
# cron_rm(id = "lotr_pipeline_daily")
```


### Example cron schedule expressions

```{r eval = FALSE}
# Every 5 minutes
cron_add(command = cron_rscript("script.R"), frequency = "*/5 * * * *")

# Every hour
cron_add(command = cron_rscript("script.R"), frequency = "hourly")

# Every day at 3:30 PM
cron_add(command = cron_rscript("script.R"), frequency = "daily", at = "15:30")

# Every Monday at 9 AM
cron_add(command = cron_rscript("script.R"), frequency = "0 9 * * 1")

# First day of every month
cron_add(command = cron_rscript("script.R"), frequency = "0 9 1 * *")
```

## Scheduling with `taskscheduleR` on Windows ⏰

```{r eval = FALSE}
# install.packages("taskscheduleR")
library(taskscheduleR)

# Define the path to your Master R script
makefile_path <- here::here("03_automation","lotr_project","scripts","Makefile.R")

# Schedule our LOTR pipeline to run daily at 8 AM
taskscheduler_create(
  taskname = "daily",
  rscript = makefile_path,
  schedule = "DAILY",
  starttime = "08:00"
)

# List all scheduled tasks
taskscheduler_ls()

# To remove a scheduled task
# taskscheduler_delete(taskname = "lotr_pipeline_daily")
```


### Example `taskscheduleR` schedule expressions

```{r eval = FALSE}
# Every 5 minutes
taskscheduler_create(taskname = "lotr_pipeline_5min", rscript = "path/to/your/Makefile.R", schedule = "MINUTE", modifier = 5)

# Every hour
taskscheduler_create(taskname = "lotr_pipeline_hourly", rscript = "path/to/your/Makefile.R", schedule = "HOURLY")

# Every day at 3:30 PM
taskscheduler_create(taskname = "lotr_pipeline_daily_1530", rscript = "path/to/your/Makefile.R", schedule = "DAILY", starttime = "15:30")

# Every Monday at 9 AM
taskscheduler_create(taskname = "lotr_pipeline_weekly", rscript = "path/to/your/Makefile.R", schedule = "WEEKLY", starttime = "09:00", days = "MON")

# First day of every month
taskscheduler_create(taskname = "lotr_pipeline_monthly", rscript = "path/to/your/Makefile.R", schedule = "MONTHLY", starttime = "09:00", days = 1)
```



# Acknowledgements {-}

This script was drafted by [Tom Arendt](https://github.com/tom-arend) and [Lisa Oswald](https://lfoswald.github.io/), with contributions by [Steve Kerr](https://smkerr.github.io/), [Hiba Ahmad](https://github.com/hiba-ahmad), [Carmen Garro](https://github.com/cgarroca), [Sebastian Ramirez-Ruiz](https://seramirezruiz.github.io/), [Killian Conyngham](https://github.com/Killian-Conyngham), [Carol Sobral](https://github.com/cbsobral), [Laia Domenech Burin](https://ldmnch.github.io/layitx/about.html), and [Linus Hageman](https://linushagemann.de).

The Lord of the Rings dialogue analysis example is based on an the Github for STAT545 at UBC. You can find their version [here](https://github.com/STAT545-UBC/STAT545-UBC-original-website/blob/master/automation01_slides/slides.md).