---
title: "Hackathon and Review"
author: "Killian Conyngham and Carol Sobral"
date: "(Fall 2025) Introduction to Data Science"
output: 
    rmdformats::robobook:
    toc: TRUE
    df_print: paged
    number_sections: FALSE
    highlight: tango
    toc_depth: 3
    toc_float: true
    self_contained: false
---

```{=html}
<style>
.h1,h2,h3 {
color:#2f1a61;
}

.subtitle, section.normal {
color:#291854;
}

.title {
color:#cc0065;
}

.nav-pills>li>a{
color: #2f1a61;
}

.nav-pills>li.active>a, .nav-pills>li.active>a:hover, .nav-pills>li.active>a:focus {
color: #fff;
background-color: #2f1a61;
}

.nav-tabs>li>a{
color: #2f1a61;
}

.nav-tabs>li.active>a, .nav-tabs>li.active>a:hover, .nav-tabs>li.active>a:focus {
color: #fff;
background-color: #2f1a61;
}

table {
  margin-left: 0;
}

</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This session is designed to help you prepare for the hackathon. We'll start with Q&A about the tasks, then review the some of the skills you might need.

------------------------------------------------------------------------

# Part 1: Hackathon Q&A


------------------------------------------------------------------------

# Part 2: Skills Review


```{r include=FALSE}
library(tidyverse)
library(palmerpenguins)
```

------------------------------------------------------------------------

## Data wrangling with dplyr

The dplyr verbs you'll use constantly:

```r
# select: choose columns
penguins |> select(species, bill_length_mm, body_mass_g)

# filter: choose rows based on conditions
penguins |> filter(species == "Adelie", year == 2007)

# mutate: create or modify columns
penguins |> mutate(body_mass_kg = body_mass_g / 1000)

# group_by + summarize: aggregate data
penguins |> 
  group_by(species) |> 
  summarize(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    n = n()
  )

# arrange: sort rows
penguins |> arrange(desc(body_mass_g))
```

Other useful functions:

```r
# count: shortcut for group_by + summarize(n = n())
penguins |> count(species, island)

# case_when: vectorized if-else for creating categories
penguins |> 
  mutate(
    size = case_when(
      body_mass_g < 3500 ~ "small",
      body_mass_g < 5000 ~ "medium",
      TRUE ~ "large"
    )
  )

# across: apply a function to multiple columns
penguins |> 
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

```


Reshaping: pivot_longer -- turn many columns into key-value pairs. pivot_wider -- spread key-value pairs into multiple columns.

```{r}
# pivot_longer
penguins_long <- penguins |>
  select(species, starts_with("bill_")) |>
  mutate(id = row_number()) |>
  pivot_longer(
    cols = starts_with("bill_"),
    names_to = "measure",
    values_to = "value"
  )
head(penguins_long)


# pivot_wider
penguins_long |>
  pivot_wider(
    names_from = measure,
    values_from = value
  ) |> 
  head()
```


------------------------------------------------------------------------

## Writing functions

Recall the basic structure of a function:

```r
function_name <- function(arguments) {
  # body: what the function does
  # last expression is returned (or use return())
}
```

A simple example:

```{r}
calculate_stats <- function(x) {
  c(
    mean = mean(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    n = length(x)
  )
}

calculate_stats(penguins$bill_length_mm)
```


### Input validation

Good functions check their inputs and fail with helpful messages:

```
my_function <- function(data, var_name) {
  
  # Check data type
  if (!is.data.frame(data)) {
    stop("'data' must be a data frame")
  }
  
  # Check if column exists
  if (!var_name %in% names(data)) {
    stop(paste("Column", var_name, "not found in data"))
  }
  
  # Check if variable is the right type
  
  # ... rest of function
}
```

### Passing column names to functions

Two main approaches in the tidyverse:

**Approach 1: Column name as string + `.data[[]]`**

```{r}
group_mean <- function(data, group_col, value_col) {
  data |> 
    group_by(.data[[group_col]]) |> 
    summarize(mean_value = mean(.data[[value_col]], na.rm = TRUE))
}

group_mean(penguins, "species", "bill_length_mm")
```

**Approach 2: Bare column name + `{{ }}`** (tidy evaluation)

```{r}
group_mean2 <- function(data, group_col, value_col) {
  data |> 
    group_by({{ group_col }}) |> 
    summarize(mean_value = mean({{ value_col }}, na.rm = TRUE))
}

group_mean2(penguins, species, bill_length_mm)
```

------------------------------------------------------------------------


## Iteration with purrr

The `map()` family applies a function to each element of a list or vector:

- `map()` returns a list
- `map_dbl()` returns a numeric vector
- `map_chr()` returns a character vector
- `map_dfr()` returns a data frame (row-binding results)

Three ways to pass functions to `map()`:

```r
# 1. Named function
map(penguins |> select(where(is.numeric)), mean, na.rm = TRUE)

# 2. Anonymous function
map(penguins |> select(where(is.numeric)), \(x) mean(x, na.rm = TRUE))

# 3. Formula shorthand (older style)
map(penguins |> select(where(is.numeric)), ~ mean(.x, na.rm = TRUE))
```

Example: applying a custom function across multiple variables. 

```{r}
# Define the function first
get_stats <- function(var) {
  data.frame(
    variable = var,
    mean = mean(penguins[[var]], na.rm = TRUE),
    sd = sd(penguins[[var]], na.rm = TRUE)
  )
}

# Then map it
numeric_vars <- c("bill_length_mm", "bill_depth_mm", "flipper_length_mm")
map_dfr(numeric_vars, get_stats)
```

------------------------------------------------------------------------

## Data visualization with ggplot2

The basic structure of a ggplot:

```r
ggplot(data, aes(x = ..., y = ..., color = ..., fill = ...)) +
  geom_...() +
  scale_...() +
  labs(title = ..., x = ..., y = ..., caption = ...) +
  theme_...() +
  theme(...)
```
### Common plot types

```{r}
library(gridExtra)

# histogram
p1 <- ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram(binwidth = 2) +
  labs(
  title = "Histogram",
  subtitle = "Use for the distribution of a numeric variable",
  x = "Bill length (mm)",
  y = "Count"
  ) +
  theme_minimal(base_size = 10)


# boxplot
p2 <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(alpha = 0.8) +
  labs(
  title = "Boxplot",
  subtitle = "Use to compare distributions across categories",
  x = "Species",
  y = "Body mass (g)"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")

# scatterplot
p3 <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(alpha = 0.7) +
  labs(
  title = "Scatterplot",
  subtitle = "Use for relationships between two numeric variables",
  x = "Flipper length (mm)",
  y = "Body mass (g)"
  ) +
  theme_minimal(base_size = 10)

# another scatterplot
p4 <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  labs(
  title = "Scatterplot by Group",
  subtitle = "Use when comparing numeric relationships across \ngroups",
  x = "Flipper length (mm)",
  y = "Body mass (g)",
  color = "Species"
  ) +
  theme_minimal(base_size = 10)


grid.arrange(p1, p2, p3, p4, ncol = 2)
```


### Faceting

Another way to show groups is to split your plot into panels. An example:

```{r}
ggplot(penguins, aes(x = bill_length_mm, fill = species)) +
  geom_histogram(binwidth = 2, color = "white") +
  labs(
    title = "Faceted Histogram",
    subtitle = "Use to compare the distribution of a numeric variable across groups",
    x = "Bill length (mm)",
    y = "Count"
  ) +
  facet_wrap(~ species, ncol = 1) +
  theme_minimal() +
  theme(legend.position = "none")
```


### Making plots standalone

A good plot should be understandable without reading surrounding text. Be mindful of labels, titles, legends, and captions.

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Penguin Body Mass vs. Flipper Length",
    subtitle = "Data from Palmer Station, Antarctica (2007-2009)",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species",
    caption = "Source: palmerpenguins  package"
  ) +
  theme_minimal()
```


## Tables

Sometimes you need a table to show the information clearly. 

```{r}
library(gt)

# summarize data
penguin_summary <- penguins |>
  group_by(species) |>
  summarize(
    mean_mass_g = round(mean(body_mass_g, na.rm = TRUE)),
    mean_flipper_mm = round(mean(flipper_length_mm, na.rm = TRUE)),
    n = n()
  )

# table
penguin_summary |> 
  gt() |> 
  tab_header(
    title = "Summary of Penguins",
  ) |> 
  cols_label(
    species = "Species",
    mean_mass_g = "Mean Mass (g)",
    mean_flipper_mm = "Mean Flipper Length (mm)",
    n = "Count"
  ) |> 
  tab_style(
    style = cell_fill(color = ),
    locations = cells_body(rows = species == "Adelie")
  )
```



------------------------------------------------------------------------

## Modeling with broom and modelsummary

### Running models

```{r}
model <- lm(body_mass_g ~ flipper_length_mm + species, data = penguins)
summary(model)
```

### Tidying with broom

```{r}
library(broom)

# coefficients as a tidy data frame
tidy(model)

# model fit statistics
glance(model)

# add fitted values and residuals to data
augment(model) |> head()
```

### Presenting with modelsummary

```{r}
library(modelsummary)

# fit multiple models
m1 <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
m2 <- lm(body_mass_g ~ flipper_length_mm + species, data = penguins)
m3 <- lm(body_mass_g ~ flipper_length_mm * species, data = penguins)

models <- list(
  "Flipper only" = m1,
  "+ Species" = m2,
  "Interaction" = m3
)

# regression table
modelsummary(
  models,
  gof_omit = "AIC|BIC|Log.Lik|RMSE",
  output = "gt"
) |> 
  tab_options(table.align = "center")
```



```{r}
# coefficients plot
modelplot(models, coef_omit = "Intercept") +
  theme_minimal() +
  labs(title = "Coefficients")
```



------------------------------------------------------------------------

This script was drafted by [Killian Conyngham](https://github.com/Killian-Conyngham) and [Carol Sobral](https://github.com/cbsobral).
